---
output: github_document
bibliography: references.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# GTclust : A package for fast clustering of spatial or temporal data with contiguity constrained hierarchical clustering

<!-- badges: start -->
<!-- badges: end -->

GTclust builds on top of `?gtclust_graph` it's main function to offers fast clustering of spatial or temporal data with contiguity constrained hierarchical clustering. `?gtclust_graph` is a quite classical hierarchical clustering but which is designed to takes advantage of contiguity constraints defined with a graph between data-points. The contiguity naturally create a sparsely connected graph that can be leveraged to speed-up the calculations, thanks to efficient data-structure [@Ambroise2019], from $O(N^2D)$ to $O(M(log(M)+D))$, with $M$ the number of links in the contiguity graph. This allow to process 10^5 data-points in seconds. To reach these performances, the dissimilarity matrix can't be computed and the algorithm resort to a more simple "stored data" approach [@Murtagh2012], which even if known to be less efficient in the general case are well fitted for the contiguity constrained problems. This approach allows to have a low spatial complexity of $O(N)$ but is less efficient with `single` and `complete` linkage criterions. Furthermore, this make the use of non-euclidean dissimilarity measures impossible. Still one may used the classical linkage criterion compatible with a storage based approach and available in `hclust` or `agnes` : 

- `ward` minimum within-cluster variance 
- `centroid` or WPGMC
- `median` or UPGMC

Furthermore, GTclust also offers two Bayesian linkage criterion [@Heller2005] that enable model selection :  

- `bayes_mom` mixture of Multinomial for counts data
- `bayes_dgmm` diagonal Gaussian mixture models for continuous features

To ease, the contiguity graph creation process, gtclust offers several interfaces to easily works with geographical, temporal (the gt in GTclust comes from here) or sequential data:

- `?gtclust_temp` to cluster sequential data, the contiguity graph follow from the data ordering 
- `?gtclust_poly` to cluster data associated to geographical polygons, the contiguity graph follow from shared boundaries
- `?gtclust_delaunay` to cluster data associated to geographical points, the contiguity graph is derived from the Delaunay triangulation of the points (thanks to the `? RTriangle` package [@RTriangle])
- `?gtclust_knn` to cluster data associated to geographical points, the contiguity graph is derived from the symmetrized knn graph of the geographical points (thanks to the `?RANN` package [@RANN])
- `?gtclust_dist` to cluster data associated to geographical points, the contiguity graph is derived from a threshold over distance the geographical points 

It also offers several methods dedicated ease the manipulations of results with spatial data thanks to a tight integration with the sf package [@Pebesma2018].



## Installation

You can install the development version of gtclust from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("comeetie/gtclust")
```

## Example

This is a basic example, we first prepare some spatial polygons data, here we used modes share in several municipalities around Paris.

```{r example-data,message=FALSE}
library(gtclust)
library(dplyr)
library(sf)
library(ggplot2)
library(ggpubr)
data("modesshare.idf")

# compute row sums
modesshare.idf <- modesshare.idf |> 
  rowwise(CODE_IRIS) |> 
  mutate(total = sum(c_across(nodep:tcom)))

# normalize per rows
modesshare.idf.percent = modesshare.idf |> 
  filter(total!=0) |>
  transmute(across(nodep:tcom,\(v){v/total})) 
  
```


To do the clustering, we use the `poly` flavor of GTclust and just provide the polygons data.frame we had just prepared. Then we may use the classical function from `?hclust`, `?cutree` to cut the dendrogram at a specific level of the `?plot.gtclust` method to draw the dendrogram:

```{r clustering,fig.show='hold',out.width="100%",fig.width=12,fig.height=6}
hc=gtclust_poly(modesshare.idf.percent,method="ward")
plot(hc)+
  ggtitle("Dendrogram of the modesshare.idf dataset","with ward linkage")+
  scale_y_continuous("Within-cluster intertia")
cutree(hc,k=30) |> head(20)
```


You may also use the `?geocutree` function which build directly a spatial data.frame with the clustering results: 
```{r geoagg}
modesshare_agg = geocutree(hc,k=500)
```

```{r,echo=FALSE}
theme_map <- function(...) {
  theme_minimal() +
  theme(
    text = element_text( color = "#22211d"),
    axis.line = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(fill = "#ffffff", color = NA), 
    panel.background = element_rect(fill = "#ffffff", color = NA), 
    legend.background = element_rect(fill = "#ffffff", color = NA),
    panel.border = element_blank(),
    ...
  )
}
```

```{r plot,fig.show='hold',out.width="100%",fig.width=14,fig.height=5,echo=FALSE}

magg = ggplot(modesshare_agg)+
  geom_sf(aes(fill=voiture*100),size=0)+
  theme_map()+scale_fill_distiller("Car (%) :",palette = "Reds",direction = 1,limits=c(0,100))+
  ggtitle(paste0("Clustering with ",nrow(modesshare_agg)," clusters"))

mraw = ggplot(modesshare.idf.percent)+
  geom_sf(aes(fill=voiture*100),size=0)+
  theme_map()+
  scale_fill_distiller("Car (%) :",palette = "Reds",direction = 1,limits=c(0,100))+
  ggtitle(paste0("Raw data with ",nrow(modesshare.idf)," polygons"))

ggarrange(mraw,magg,nrow = 1,common.legend = TRUE,legend="bottom")
```



```{r plotparis,fig.show='hold',out.width="100%",fig.width=14,fig.height=5,echo=FALSE}
modesshare.paris = modesshare.idf.percent|>filter(substr(CODE_IRIS,1,2)==75)
env_paris=st_union(modesshare.paris) |> st_geometry()
modesshare_agg_paris = st_intersection(modesshare_agg,env_paris)

magg_paris = ggplot(modesshare_agg_paris)+
  geom_sf(aes(fill=tcom*100),size=0)+
  theme_map()+
  scale_fill_distiller("Transit (%) :",palette = "Greens",direction = 1,limits=c(0,100))+
  ggtitle(paste0("Clustering with ",nrow(modesshare_agg_paris)," polygons"))

mraw_paris = ggplot(modesshare.paris)+
  geom_sf(aes(fill=tcom*100),size=0)+
  theme_map()+
  scale_fill_distiller("Transit (%) :",palette = "Greens",direction = 1,limits=c(0,100))+
  ggtitle(paste0("Raw data with ",nrow(modesshare.paris)," polygons"))

ggarrange(mraw_paris,magg_paris,nrow = 1,common.legend = TRUE,legend="bottom")

```


```{r,fig.show='hold',out.width="100%",fig.width=10,fig.height=5,warning=FALSE}
res_mom = gtclust_delaunay(modesshare.idf |> st_centroid(),method="bayes_mom")
ggplot(data.frame(ts=res_mom$test.stat,K=(nrow(res_mom$data)-1):1))+
    geom_point(aes(x=K,y=ts))+theme_bw()+scale_y_continuous("Test statistic")+ggtitle("Bayesian test score","with respect to the number of cluster")
```

```{r benchplot,fig.show='hold',out.width="100%",fig.width=10,fig.height=5,echo=FALSE}
data("res.bench")

ggplot(res.bench|>mutate(time=time*10^-9))+
  geom_line(aes(x=N,y=time,group=alg,color=alg))+
  geom_point(aes(x=N,y=time,group=alg,color=alg))+
  geom_text(data=res.bench|>mutate(time=time*10^-9)|>filter(N==max(N)),aes(x=N,y=time,label=alg,color=alg),nudge_x = 500,hjust = "left",vjust="midlle")+theme_bw()+scale_x_continuous(expand = c(0.2,1.5))+scale_color_discrete(guide ="none")+scale_y_continuous("Time (s)")+ggtitle("Running times","of graph building and gtclust variants.")

```

# References